---
title: "Lecture 3: Data Quality "
output: html_document
---

## Lecture 3: Data Quality

The quality of data can be a selling point or achilles heel of a data science project.

Checking data quality

The objective of this lecture is to handle missing values appropriately and script visual checks to find errors introduced in data input/output. We will also start to view computational optimization techniques, like taking advantage of multiple cores for heavy duty operations (parallel processing).

##### Create a random dataset

Consider a dataset of *n* individuals froma  community
```{r}

n = 10000

set.seed(123)
df <- data.frame(id=seq(1,n,1), 
                
                age = sample(21:65,n, replace=T), 
                education = sample(c("High School","High School","High School","College","College","Graduate",NA),n,replace=T),
                gender = sample(c("M","F","-999"),n,replace=T)
                )

```

##### Examine dataset
```{r}
summary(df)
str(df)
```


##### 
```{r}
table(df$education)
table(df$education, useNA = "always")
```


###Exercise 3.1
GHCN-M: Missing values analysis in matrix of weather anomalies from 1880 to Present


Get data: [NOAA GHCN-M](https://www.kaggle.com/noaa/global-historical-climatology-network)

```{r}

```

Weather: Matching human-reported weather events to radar events over time and space

